A Collection of Thoughts on the Keccak/SHA-3 Construction

### 2018-04-29 s2018-math490/thesis/thesis.tex
Update thesis
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

To construct the inverse of $\theta$, note that
$A'[x, y, z] = A[x, y, z] \oplus D[x, z]$; hence,
$A[x, y, z] = A'[x, y, z] \oplus D'[x, z]$ for some $D' = D$. Since $D[x, z]$
is composed of several $C[x, z]$, where
$C[x, z] = \oplus_{y = 0}^{4} A[x, y, z]$, we can similarly define $C'[x, z]$
to be $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$. Then, we expect the inverse
of $\theta$ to be of a similar form. This reduces to a linear algebra problem
over boolean variables. We know that $D'[x, z] = D[x, z]$ in order to recover
$A[x, y, z]$. Hence we can represent $D[x, z]$ as a series of bits of length
$5 \times w$, where bit $i = z' + 5 \times x'$ is $1$ if and only if
$C[x', z']$ is used in the construction of $D[x, z]$. Further, we can view each
of the $C'[x, z]$ as being the conjunction of three $C[x', z']$ in $A$, and thus
can represent these as bit strings where bit $j = z' + 5\times x'$ is 1 if
and only if $C[x', z']$ is used to construct $C'[x, z]$. (That is,
since $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$,
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus D[x, z])$ and hence
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus C[x', z'] \oplus C[x'', z''])$,
and thus $C'[x, z] = C[x, z] \oplus C[x', z'] \oplus C[x'', z'']$, for some
$x, z, x', z', x'', z''$ based on the definition of $\theta$. Thus,
giving each $C'[x, z]$ a constant $c_{x,z}$ for whether it is used in constructing
$D'[x, z]$, we can form a system of linear equations and solve for the
constants $c_{x, z}$ in each expression. Since there are $5 \times w$ variables
and $5 \times w$ equations in each equation for $D'[x, z]$, this can be solved
easily, yielding the inverse of $\theta$.

@@

Overall, Keccak/SHA-3 remains secure to many attacks from SAT solvers.

>>>>

To construct the inverse of $\theta$, note that
$A'[x, y, z] = A[x, y, z] \oplus D[x, z]$; hence,
$A[x, y, z] = A'[x, y, z] \oplus D'[x, z]$ for some $D' = D$. Since $D[x, z]$
is composed of several $C[x, z]$, where
$C[x, z] = \oplus_{y = 0}^{4} A[x, y, z]$, we can similarly define $C'[x, z]$
to be $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$. Then, we expect the inverse
of $\theta$ to be of a similar form. This reduces to a linear algebra problem
over boolean variables. We know that $D'[x, z] = D[x, z]$ in order to recover
$A[x, y, z]$. Hence we can represent $D[x, z]$ as strings of length
$5 \times w$, where bit $i = z' + 5 \times x'$ is $1$ if and only if
$C[x', z']$ is used in the construction of $D[x, z]$. Further, we can view each
of the $C'[x, z]$ as being the conjunction of three $C[x', z']$ in $A$; thus
these are strings where bit $j = z' + 5\times x'$ is 1 if and only if
$C[x', z']$ is used to construct $C'[x, z]$.

(That is, since $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$,
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus D[x, z])$ and hence
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus C[x', z'] \oplus C[x'', z''])$,
and thus $C'[x, z] = C[x, z] \oplus C[x', z'] \oplus C[x'', z'']$, for some
$x, z, x', z', x'', z''$ based on the definition of $\theta$).

Thus, giving each $C'[x, z]$ a constant $c_{x,z}$ for whether it is used in
constructing $D'[x, z]$, we can form a system of linear equations and solve for
the constants $c_{x, z}$ in each expression. Since there are $5 \times w$
variables and $5 \times w$ equations in each equation for $D'[x, z]$, this can
be solved easily, yielding the inverse of $\theta$.

@@

Overall, Keccak/SHA-3 remains secure to many attacks from SAT solvers. The
majority of the techniques discussed here do not scale to the full $w=64$,
rendering them ineffective for attacking SHA-3 in general. More efficient
techniques are thus necessary for finding collisions in SHA-3; this likely
involves finding ways in which structure in small values of $w$ translate to
structure in larger values of $w$.

Future work includes scaling the partial fixed point attacks via intermediate
models, evaluating weaknesses in alternate constructions, and improving
the run time of marginal and differential analysis. However, without
significant breakthroughs in logical cryptanalysis, most of these attacks are
unlikely. Further, it is not immediately obvious what effect low permutation
order has on the possibility of collisions; introducing permutation-based
cryptanalysis could result in new techniques applicable to SHA-3.
